---
title: "cm007 Exercises: Practice with `dplyr`"
author: "Trevor Baker"
date: "24/09/2019"
output: 
  html_document:
    keep_md: true
    theme: paper
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

```{r}
install.packages("tsibble")
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tsibble))
```


This worksheet contains exercises aimed for practice with `dplyr`. 


1. (a) What's the minimum life expectancy for each continent and each year? (b) Add the corresponding country to the tibble, too. (c) Arrange by min life expectancy.

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(min_life = min(lifeExp),
            country = country[which(lifeExp==min_life)]) %>%
  arrange(min_life)

```

2. Calculate the growth in population since the first year on record _for each country_ by rearranging the following lines, and filling in the `FILL_THIS_IN`. Here's another convenience function for you: `dplyr::first()`. 

```
mutate(rel_growth = FILL_THIS_IN) %>% 
arrange(FILL_THIS_IN) %>% 
gapminder %>% 
DT::datatable()
group_by(country) %>% 
```

```{r}
#method 1 - using index[1]
gapminder %>%
  group_by(country) %>% 
  mutate(rel_growth = (pop-pop[1])/pop) %>% 
  arrange(year) %>% 
  DT::datatable()
  
#method 2 - using first()
gapminder %>%
  group_by(country) %>% 
  arrange(year) %>% 
  mutate(rel_growth = (pop-first(pop))/pop) %>% 
  DT::datatable()


```

```{r}

```


3. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent, sorted by the drop, by rearranging the following lines of code. Ensure there are no `NA`'s. Instead of using `lag()`, use the convenience function provided by the `tsibble` package, `tsibble::difference()`:

```
drop_na() %>% 
ungroup() %>% 
arrange(year) %>% 
filter(inc_life_exp == min(inc_life_exp)) %>% 
gapminder %>% 
mutate(inc_life_exp = FILL_THIS_IN) %>% 
arrange(inc_life_exp) %>% 
group_by(country) %>% 
group_by(continent) %>% 
knitr::kable()
```

```{r}

#method 1
gapminder %>% 
  group_by(country) %>% 
  arrange(year) %>% 
  mutate(inc_life_exp = tsibble::difference(lifeExp)) %>% 
  drop_na() %>% #have to drop NAs before we filter
  ungroup() %>% 
  group_by(continent) %>%   
  filter(inc_life_exp == min(inc_life_exp)) %>% 
  arrange(inc_life_exp) %>%
  knitr::kable()


#method 2
gapminder %>% 
  group_by(country) %>% 
  arrange(year) %>% 
  mutate(inc_life_exp = tsibble::difference(lifeExp)) %>% 
  #drop_na() %>% #have to drop NAs before we filter
  ungroup() %>% 
  group_by(continent) %>%   
  filter(inc_life_exp == min(inc_life_exp, na.rm=T)) %>% #drop NAs with na.rm=T
  arrange(inc_life_exp) %>%
  knitr::kable()


```



