---
title: "cm010_notes"
author: "Trevor Baker"
date: "03/10/2019"
output: html_document
---

```{r setup}

library(tidyverse)
library(tidyr)


```



##Tibble joins

 
### Mutating joins

- adding new column(s) onto base tibble
- see r4ds chapter 13.4

```{r join_mutate}

x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)

```

#### inner join

 - like a `merge()`
 
```{r join_inner}

inner_join(x, y, by = "key") #keep only common keys

```

#### outer join

__left_join:__ keep all obs in x; this is like the default join, adding columns to your base table x
__right_join:__ keep all obs in y; right_join(x,y) is same as left_join(y,x)
__full_join:__ keep all obs in both

```{r join_outer}

left_join(x, y) #keeps key 1,2,3
right_join(x, y) #keeps key 1,2,4
full_join(x, y) #keeps all keys

```

 
 ### filter joins
 
 __semi_join:__ keep all in x that have match in y
 __anti_join:__ drop all in x that have match in y
 
 But the _columns from y are not added_ 
 - so it's basically a way of filtering down your x table based on y's contents, rather than joining columns together.
 
 
```{r}
semi_join(x,y)
anti_join(x,y)

```

### set operations

- these are `intersect()`, `union()`, `setdiff()`
- not a way of joining but of _identifying common or differing members_, which could then be used for joining

- intersect
  + return common members
- union
  + return all members of both
- setdiff
  + return members of x are not in y
  
If you do these on a dataframe, then it evaluates by row, i.e. treats one row the way a single value is treated in a vector.

### binding

__tidyr__ has `bind_rows()` and `bind_cols()` that function like `rbind()` and `cbind()`

__bind_rows__ both tables need to have same columns, then bind together
__bind_cols__ both tables need to have same rows, then bind together
 