---
title: "cm007-notes"
author: "Trevor Baker"
date: "24/09/2019"
output: github_document
---

```{r setup, message=F, warning=F, echo=F}

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(tsibble))
suppressPackageStartupMessages(library(here))
library(DT)

#wrong because a<3 not vectorized
tibble(a = 1:4) %>% 
  mutate(b = (if (a < 3) "small" else "big"))

#two possible ways to do it
tibble(a = 1:4) %>% 
  mutate(c = sapply(a, function(x) if (x < 3) "small" else "big"),
         d = purrr::map_chr(a, ~  (if(.x < 3) "small" else "big")))


```

Diff between summarize and mutate:
```{r}

library(gapminder)
gapminder %>%
  summarize(mu = mean(lifeExp)) #one column, one row
  
gapminder %>%
  mutate(mu = mean(lifeExp)) #mu value tacked on as column, all rows still there
  

```


`group_by`
can group by a novel variable, adding a new column to the tibble:
```{r}

gapminder %>%
  group_by(small.lifeExp= lifeExp < 60)

```

Add identifying column to your group_by tibble:
```{r}

gapminder %>%
  mutate(group=paste(continent,year,sep="_")) %>%
           group_by(continent, year)

```

_Window functions_
take in vector, return a vector that depends on the entire vector
e.g. `rank()` - you can't get the right output going linearly along vector, need to look at whole
whereas `round()` just works linearly along vector


`lubridate`
```{r}

#can standardize a messy bunch of date inputs - thes emust be in month-day-year order though
ymd <- lubridate::mdy("Sep 24. 2019", 
               "09 25 19", 
               "Sept 26 2019",
               "Septem 27 2019",
               "September 28 2019")

#lubridate does not do year-month format but tsibble does
tsibble::yearmonth(ymd)


```

`tsibble`
can create tsibble object like so, give date column as "index"
```{r}

#didn't get this to work, couldnt get data loaded
# energy <- here::here("data", "daily_consumption.csv") %>% 
#   read_csv()
# energy <- as_tsibble(energy)


```




