---
title: 'Assignment 1: gapminder slides'
author: "Trevor Baker, 98610074"
date: "17/09/2019"
output:
  beamer_presentation: default
  ioslides_presentation:
    smaller: yes
---

```{r knitr_opt, include = FALSE}
knitr::opts_chunk$set(out.height = "50%",  out.width = "70%")
```

```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}

library(gapminder)
data("gapminder")

#shorten name
gm <- gapminder


```

## Basic dataset overview
'gapminder' is a dataset providing demographic data from 1952 to 2007. Its numeric variables include year, life expectancy, population and GDP per capita. Here is a simple statistical summary of them:

\tiny
```{r quick_summ, echo=FALSE}

options(scipen=8) #prevent scientific notation in population summary
summary(gm[ ,c("year", "lifeExp", "pop", "gdpPercap")]) #summarize numeric columns

```
\normalsize

## GDP ranges by continent
<font size="1"> While Oceania is the richest continent on average, Kuwait, which is on the Asian continent, is the richest country overall. In the graph below, all countries with average income values greater than $50000 have been labelled. These include Kuwait, Norway, Singapore, and the United States. </font>   
```{r boxplot, echo=FALSE}

{ 
  par(mar=c(5,6,1,1))
  boxplot(gm$gdpPercap~gm$continent,
        xlab="",
        ylab = "",
        yaxt="n",
        las=3,
        col="lightgrey")
  
  #y tick mark values
  ytick <- seq(0,120000,20000)
   
  #make y axis
  axis(2, 
       at=seq(0,120000,20000), 
       labels=format(ytick, scientific = FALSE), 
       las=1)
  mtext("GDP per capita ($ US, inflation-adjusted)", 
        side=2, #cex=0.7, 
        line=4)
  
  #text for points with high income
  thresh <- 40000
  text( x=as.factor(gm$continent[gm$gdpPercap>thresh]),
        y=gm$gdpPercap[gm$gdpPercap>thresh],
        labels = paste(gm$country[gm$gdpPercap>thresh],gm$year[gm$gdpPercap>thresh],sep="-") ,
        cex=0.5,
        pos=4,
        offset=0.5)
  
}

```


## GDP trends by country over time 
<font size="2"> Most countries had a positive trend in GDP over the dataset period, but a few, such as the Democratic Republic of Congo, actually decreased in inflation-adjusted GDP. </font>   
```{r trends, echo=FALSE}

#select and merge data for first and last years
gm1952 <- plyr::ddply(gm[gm$year==1952,], ~country, plyr::summarise,
                      inc1952=gdpPercap,
                      continent=continent)
gm2007 <- plyr::ddply(gm[gm$year==2007,], ~country, plyr::summarise, 
                      inc2007=gdpPercap,
                      continent=continent)
gm2 <- merge(gm1952, gm2007)

#calculate and sort by % diff in GDP over full dataset duration
gm2$pcdiff <- round((gm2$inc2007-gm2$inc1952)/gm2$inc1952 * 100, 1)
gm2 <- gm2[order(gm2$pcdiff, decreasing = T),]

big.incr <- as.character( gm2$country[1] ) #which country had largest increase?
big.decr <- as.character( gm2$country[nrow(gm2)] ) #which country had largest decrease?

{

  plot(x=gm$year[gm$country==big.incr],
       y=gm$gdpPercap[gm$country==big.incr],
       ylim=c(0,120000),
       bg="lightgreen",
       pch=21,
       type="b",
       main="GDP trends for selected countries, 1952 to 2007",
       xlab="Year",
       ylab="GDP per capita ($ US,inflation-adjusted)")
  points(x=gm$year[gm$country==big.decr],
         y=gm$gdpPercap[gm$country==big.decr],
         bg="red",
         type="b",
         pch=21)
  points(x=gm$year[gm$country=="Canada"],
         y=gm$gdpPercap[gm$country=="Canada"],
         bg="black",
         type="b",
         pch=21)
  points(x=gm$year[gm$country=="China"],
         y=gm$gdpPercap[gm$country=="China"],
         bg="yellow",
         type="b",
         pch=21)
  points(x=gm$year[gm$country=="Norway"],
         y=gm$gdpPercap[gm$country=="Norway"],
         bg="blue",
         type="b",
         pch=21)
  points(x=gm$year[gm$country=="Kuwait"],
         y=gm$gdpPercap[gm$country=="Kuwait"],
         bg="orange",
         type="b",
         pch=21)
  legend("topright",
         legend=c("Norway", "Kuwait", "Canada",big.incr,"China",big.decr),
         pch=21,
         pt.bg=c("blue","orange","black","lightgreen","yellow","red"))

  
}
```

## GDP and life expectancy
<font size="2"> You can see general groupings of GDP and life expectancy by continent in the graph below. There seems to be an asymptotic relationship between life expectancy and GDP - mean life expectancy doesn't increase much beyond 80 years, no matter how rich the country. </font>

```{r lifeExp_GDP, echo=FALSE}

leg.labels <- levels(as.factor(gm$continent))

{ 
  plot(gm$gdpPercap,
       gm$lifeExp, 
       xlim=c(0,50000),
       ylab="Life expectancy (years)",
       xlab="GDP per capita",
       main="Life expectancy vs GDP by continent",
       col=as.factor(gm$continent))
  
  legend("bottomright", pch=21, pt.bg=1:length(leg.labels),
       legend=leg.labels ) 
}

```

