---
title: "cm012 Exercises: Factors"
output: 
  html_document:
    keep_md: true
    theme: paper
---

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r load_pkg, echo = FALSE, message = FALSE, warning = FALSE}

library(gapminder)
library(tidyverse)
library(dplyr)
library(forcats)
library(ggplot2)

```


## Motivating the need for factors in R

### Activity 1: Using Factors for plotting 

**1.1** Let's look again into `gapminder` dataset and create a new cloumn, `life_level`, that contains five categories ("very high", "high","moderate", "low" and "very low") based on life expectancy in 1997. Assign categories accoring to the table below:

| Criteria          | life_level| 
|-------------      |-----------|
| less than 23      | very low |
| between 23 and 48 | low |
| between 48 and 59 | moderate |
| between 59 and 70 | high |
| more than 70      | very high |

Function `case_when()` is a tidier way to vectorise multiple `if_else()` statements. you can read more about this function [here](https://dplyr.tidyverse.org/reference/case_when.html).

```{r notFactor_noLevels}
gapminder %>% 
  filter(year == 1997) %>% 
  mutate(life_level = case_when(lifeExp < 23 ~ "very low",
                                lifeExp < 48 ~ "low",
                                lifeExp < 59 ~ "moderate",
                                lifeExp < 70 ~ "high",
                                TRUE ~ "very high")) %>%
  ggplot() + geom_boxplot(aes(x = life_level, y = gdpPercap)) +
  labs(y = "GDP per capita, $", x= "Life expectancy level, years") +
  theme_bw() 
```

Do you notice anything odd/wrong about the graph?

We can make a few observations:

- It seems that none of the countries had a "very low" life-expectancy in 1997. 

- However, since it was an option in our analysis it should be included in our plot. Right?

- Notice also how levels on x-axis are placed in the "wrong" order.

**1.2** You can correct these issues by explicitly setting the levels parameter in the call to `factor()`. Use, `drop = FALSE` to tell the plot not to drop unused levels 
```{r Factor_w_Levels}
gm <- gapminder %>%  
        mutate(life_level = factor(case_when(lifeExp < 23 ~ "very low", #you provide a logical evaluation and the if(TRUE) result given behind the ~
                                      lifeExp < 48 ~ "low", #so its syntax is [logical] ~ ifTRUE, ifFALSE
                                      lifeExp < 59 ~ "moderate",
                                      lifeExp < 70 ~ "high",
                                      TRUE ~ "very high") , #in this row, we don't evaluate anything, its just the last ELSE just directly give TRUE
                            levels = c("very low","low","moderate","high","very high")))

gm %>% ggplot(aes(x = life_level, y = gdpPercap)) + 
        geom_boxplot() +
        labs(y = "GDP per capita, $", x= "Life expectancy level, years") +
        scale_x_discrete(drop=F) +
        theme_bw() 
```

## Inspecting factors (activity 2)

In Activity 1, we created our own factors, so now let's explore what categorical variables that we have in the `gapminder` dataset.

### Exploring `gapminder$continent` (activity 2.1)

Use functions such as `str()`, `levels()`, `nlevels()` and `class()` to answer the following questions:

- what class is  `continent`(a factor or charecter)?
- How many levels? What are they?
- What integer is used to represent factor "Asia"?

```{r}

gapminder$continent %>%
  str()

gapminder$continent %>%
  levels()

gapminder$continent %>%
  nlevels()

gapminder$continent %>%
  class()


```

### Exploring `gapminder$country` (activity 2.2)

Let's explore what else we can do with factors:

Answer the following questions: 

- How many levels are there in `country`?
- Filter `gapminder` dataset by 5 countries of your choice. How many levels are in your filtered dataset?

```{r countLevel}

gapminder$country %>% nlevels() #142 levels

gm2 <- gapminder %>%
        filter(country %in% c("Canada", "Mexico","Japan","China","Thailand"))
gm2$country %>% nlevels() #still 142 levels


```

## Dropping unused levels

What if we want to get rid of some levels that are "unused" - how do we do that? 

The function `droplevels()` operates on all the factors in a data frame or on a single factor. The function `forcats::fct_drop()` operates on a factor.

```{r dropLevels }
#method 1
gm_drop_1 <- gm2 %>% 
  droplevels()

gm_drop_1$country %>% nlevels()
gm_drop_1$continent %>% nlevels() #in the first way the extra factors were dropped from continent as well

#method 2
gm_drop_2 <- gm2 %>% mutate(country=forcats::fct_drop(gm$country))

gm_drop_2$country %>% nlevels()
gm_drop_2$continent %>% nlevels() #the second way acts on country only and keeps all continent levels

```

## Changing the order of levels

Let's say we wanted to re-order the levels of a factor using a new metric - say, count().

We should first produce a frequency table as a tibble using `dplyr::count()`:

```{r}

gapminder %>%
  filter(year==1997) %>%
  count(continent)

```

The table is nice, but it would be better to visualize the data.
Factors are most useful/helpful when plotting data.
So let's first plot this:

```{r}
gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=continent)) +
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent")
```

Think about how levels are normally ordered. 
It turns out that by default, R always sorts levels in alphabetical order. 
However, it is preferable to order the levels according to some principle:

  1. Frequency/count. 
  
- Make the most common level the first and so on. Function `fct_infreq()` might be useful.
- The function `fct_rev()` will sort them in the opposite order.

For instance ,
    `  
```{r}
#in order of factor frequency
gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=fct_infreq(continent))) +
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent")

#reverse order
gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=fct_rev(fct_infreq(continent)))) + #use nested
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent")

gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=continent %>% fct_infreq() %>% fct_rev())) + #use all pipe
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent")

```

Section 9.6 of Jenny Bryan's [notes](https://stat545.com/factors-boss.html#reorder-factors) has some helpful examples.

  2. Another variable. 
  
  - For example, if we wanted to bring back our example of ordering `gapminder` countries by life expectancy, we can visualize the results using `fct_reorder()`. 

```{r}
gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=fct_reorder(continent, lifeExp))) + #by default it calculates it by median
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent") 


gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=fct_reorder(continent, lifeExp, max))) + #can use third argument to give different summary stat to order by
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent") 

gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=continent %>% fct_reorder(lifeExp, max))) + #same as above in pipe form
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent") 


```

Use `fct_reorder2()` when you have a line chart of a quantitative x against another quantitative y and your factor provides the color. 

```{r}
## order by life expectancy 
ggplot(gm2, aes(x = year, y = lifeExp,
                  color = fct_reorder2(country,lifeExp,year))) +
  geom_line() +
  labs(color = "country")
```

## Change order of the levels manually

This might be useful if you are preparing a report for say, the state of affairs in Oceania.

```{r relevel}
gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=continent %>% fct_relevel( c("Oceania", "Europe") ))) + #put Oceania first, don't need to specify others, they are filled by default
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent") 

gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=continent %>% 
                 fct_relevel("Oceania") %>% #this would put Oceania first
                 fct_rev() )) + #use the fct_rev to put Oceania to the last instead
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent") 

```
More details on reordering factor levels by hand can be found [here] https://forcats.tidyverse.org/reference/fct_relevel.html

### Recoding factors

Sometimes you want to specify what the levels of a factor should be.
For instance, if you had levels called "blk" and "brwn", you would rather they be called "Black" and "Brown" - this is called recoding.
Lets recode `Oceania` and the `Americas` in the graph above as abbreviations `OCN` and `AME` respectively using the function `fct_recode()`.

```{r recode}

gapminder %>%
  filter(year==1997) %>%
  ggplot() +
  geom_bar(aes(x=continent %>% 
                 fct_recode("OCN"="Oceania") )) + #use syntax "new"="old"
  coord_flip() +
  theme_bw() +
  labs(y="Number of entries", x="Continent") 


```

## Grow a factor (OPTIONAL)

Let’s create two data frames,`df1` and `df2` each with data from two countries, dropping unused factor levels.
```{r create_dfs}

df1 <- gapminder %>%
        filter(country %in% c("Canada","Japan")) %>%
        droplevels()

df2 <- gapminder %>%
        filter(country %in% c("Norway","Sweden")) %>%
        droplevels()

```

The country factors in df1 and df2 have different levels.
Can we just combine them?
```{r rbind_factors, warning=F, message=F}

df3 <- rbind.data.frame(df1,df2)

levels(df3$country) #seems that I can just combine them with rbind.data.frame

df4 <- full_join(df1,df2)
levels(df4$country)  #if using full_join, then factor column becomes character so would need to do this:
df5 <- full_join(df1,df2) %>% mutate(country=as.factor(country))
levels(df5$country)

```

The country factors in `df1` and `df2` have different levels.
Can you just combine them using `c()`?
```{r combine_w_c, warning=F, message=F}

c(df1$country, df2$country) #cannot do this, it gives integers and level 1 has one value in df1 and another in df2

```

Explore how different forms of row binding work behave here, in terms of the country variable in the result.
```{r diff_binding, warning=F, message=F}

df6 <- bind_rows(df1,df2)
df6$country %>% levels()


```
